---
title: "Monthly Toronto shelter usage in 2020"
author: 
  - Rohan Alexander
thanks: "Code and data are available at: https://github.com/RohanAlexander/starter_folder-main-2."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Shelter usage is an important factor in determining the livability of Toronto because of the large homeslessness population. We obtain monthly shelter usage in Toronto from the City on a demographic basis and analyze it using a graph. We find increased shelter usage in Winter, compared to summer and for younger age-groups compared with older ones. Our findings have implications for shelter constuction."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
```

# Introduction

First paragraph is going to be motivational and broad. 

Second paragraph is about what was done and what was found.

Third paragrph about implications.

The remainder of this paper is: Section \@ref(data) explains the data. Section \@ref(results) covers results...... 



# Data

Paragraph or two introducing the dataset broadly.

Then show an extract of the dataset (Table \@ref(tab:dataextract)).

```{r dataextract}
library(tidyverse)
library(knitr)
monthly_shelter_usage <- 
  read_csv(here::here("inputs/data/monthly_shelter_usage.csv"))

monthly_shelter_usage |> 
  select(population_group, returned_from_housing) |>
  slice(1:10) |> 
  kable(
    caption = "First ten rows of a dataset that shows shelter usage",
    col.names = c("Population group", "Number returned from housing"),
    digits = 1,
    booktabs = TRUE,
    linesep = ""
  )
```


Paragraph or two more about Table  \@ref(tab:dataextract).


We are interested in the relationship between teh number of homessless who return from houseing and the return to shelter, because this is a measure of how homeless individuals are cycling between options and has implications for public policy.

Figure X shows the relationship between the number of people reutrning from housing and the numebr of people returning to shelters.




Our data is of homessless (Figure \@ref(fig:bills)).

```{r bills, fig.cap="Bills of penguins", echo = FALSE}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = FALSE) +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

Also bills and their average (Figure \@ref(fig:billssssss)). (Notice how you can change the height and width so they don't take the whole page?)

```{r billssssss, fig.cap="More bills of penguins", echo = FALSE, fig.width=8, fig.height=4}

# This needs to be about the random data tha tI sasmpled?
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = FALSE) +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk way more about it. 



# Model

\begin{equation}
Pr(\theta | y) = \frac{Pr(y | \theta) Pr(\theta)}{Pr(y)}  (\#eq:bayes)
\end{equation}

Equation \@ref(eq:bayes) seems useful, eh?

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by @thereferencecanbewhatever If we were interested in baseball data then @citeLahman could be useful. 

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

# Discussion

## First discussion point

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional details


\newpage


# References


