---
title: "Monthly Toronto shelter usage in 2020"
author: 
  - Rohan Alexander
thanks: "Code and data are available at: https://github.com/RohanAlexander/starter_folder-main-2."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Shelter usage is an important factor in determining the livability of Toronto because of the large homeslessness population. We obtain monthly shelter usage in Toronto from the City on a demographic basis and analyze it using a graph. We find increased shelter usage in Winter, compared to summer and for younger age-groups compared with older ones. Our findings have implications for shelter constuction."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
```

# Introduction

First paragraph is going to be motivational and broad. 

Second paragraph is about what was done and what was found.

Third paragrph about implications.

The remainder of this paper is: Section \@ref(data) explains the data. Section \@ref(results) covers results...... 



# Data

Paragraph or two introducing the dataset broadly. We obtain our dataset from the City of Tornot open Data Portal, using the `opendatatoronto` package [@citeopendatatoronto] and the statistical programming lnaugage R [@citeR].

Then show an extract of the dataset (Table \@ref(tab:dataextract)).

```{r dataextract, echo = FALSE, warning=FALSE, message = FALSE}
library(tidyverse)
library(knitr)
monthly_shelter_usage <- 
  read_csv(here::here("inputs/data/monthly_shelter_usage.csv"))

monthly_shelter_usage |> 
  select(population_group, returned_from_housing) |>
  slice(1:10) |> 
  kable(
    caption = "First ten rows of a dataset that shows shelter usage",
    col.names = c("Population group", "Number returned from housing"),
    digits = 1,
    booktabs = TRUE,
    linesep = ""
  )
```


Paragraph or two more about Table  \@ref(tab:dataextract).


We are interested in the relationship between teh number of homessless who return from houseing and the return to shelter, because this is a measure of how homeless individuals are cycling between options and has implications for public policy.

Figure \@ref(fig:bills) shows the relationship between the number of people reutrning from housing and the numebr of people returning to shelters.


```{r bills, fig.cap = "Bills of penguins", echo = FALSE, warning=FALSE, message = FALSE}
monthly_shelter_usage |>
  ggplot(aes(x = returned_from_housing, 
             y = returned_to_shelter,
             color = population_group)) +
  geom_point(alpha = 0.8) +
  scale_colour_viridis_d()
```

Talk more about it.

Also bills and their average (Figure \@ref(fig:billssssss)). (Notice how you can change the height and width so they don't take the whole page?)


\newpage

\appendix

# Appendix {-}


# Additional details


\newpage


# References


